---
title: "Understanding the Impact of Rural Female Literacy on Key Socio-Health Indicators in India"
author: "Aleena Santhosh"

output: 
  pdf_document:
    keep_tex: true

editor_options: 
  chunk_output_type: console

header-includes:
  - "\\usepackage{fvextra}"
  - "\\DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\\\\{\\}}"
  - \usepackage{hyperref}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
install.packages("tinytex")
tinytex::install_tinytex(force = TRUE)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

#Load Packages
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse,
               plotly,
               janitor,
               visdat,
               esquisse, here, dplyr, ggplot2)

nfhs_data <- read_csv(here::here("data/nfhs.csv"))
```

## Introduction

Female Literacy is a key indicator of social and economic development of women in a country. Understanding its impact on varied aspects of women's lives is crucial for policy planning, gender equality, and rural development. This analysis focuses on rural female literacy rates across Indian states. The objective is to explore patterns, identify disparities, and understand how literacy impacts socio-health indicators.

The analysis uses the National Family and Health Survey-5 dataset. It seeks to explore the relationship between rural female literacy and factors like Infant Mortality Rate, Fertility Rate, and access to birth control for women.

The analysis involves data cleaning, filtering, summary statistics, and visualization.

## Data Cleaning

The National Family and Health Survey dataset contained a vast set of variables, from which the required variables for the analysis were to be filtered and cleaned. The dataset also contained information on both states and union territories. Given the scope of the analysis, only the states data was retained. All values were converted into numeric to ensure that the input for statistical analysis was accurate.

```{r}
female_lit <- 
nfhs_data %>% mutate(Female_Literacy = as.numeric(
gsub("\\(|\\)", "", `Women (age 15-49) who are literate4 (%)`)),
Infant_Mortality_Rate = as.numeric(gsub("\\(|\\)", "", 
`Infant mortality rate (per 1000 live births)`)),
`Total Fertility Rate (number of children per woman)` = as.numeric(gsub("[^0-9.]", "", 
      `Total Fertility Rate (number of children per woman)`))) %>% 
select(`States/UTs`, `Area`, `Female_Literacy`,
`Total Fertility Rate (number of children per woman)`,
`Infant_Mortality_Rate`,
`Current Use of Family Planning Methods (Currently Married Women Age 15-49  years) - Any modern method6 (%)`) %>% 
rename("Use of Modern Family Planning Methods (%)" = 
`Current Use of Family Planning Methods (Currently Married Women Age 15-49  years) - Any modern method6 (%)`,
           "States" = `States/UTs`) %>%
    filter(!States %in% c("India",
           "Andaman & Nicobar Islands",
         "Dadra and Nagar Haveli & Daman and Diu",
         "Ladakh",
         "Lakshadweep",
         "NCT of Delhi",
         "Puducherry"),
          Area == "Rural")
```

The raw dataset contained missing and inconsistent values. Some cells used the symbol `*` to represent missing data, and certain numeric columns were stored as character values.

To ensure accurate analysis, the dataset was cleaned by replacing invalid entries and converting variables into appropriate data types.

```{r}
female_lit <- female_lit %>% 
  mutate(across(
    c(`Total Fertility Rate (number of children per woman)`,
      `Use of Modern Family Planning Methods (%)`),
    ~ as.numeric(na_if(as.character(.x), "*"))
  ))
```

## Summary Statistics

After cleaning the dataset and filtering out Union Territories, summary statistics were computed to understand overall literacy patterns.

The mean rural female literacy rate provides a general measure of literacy levels across states.

```{r}
mean(female_lit$Female_Literacy, na.rm = TRUE)
```

The average rural female literacy rate across states is approximately **75.18%**. This indicates the overall level of literacy among women in rural areas, though significant variation exists between states.

```{r}
summary(female_lit$Female_Literacy)
```

The summary statistics show the minimum, maximum, median, and quartile distribution of literacy rates, highlighting disparities across states.

## Visualization

To better understand the variation in rural female literacy across states, a bar chart was created. This visualization helps compare literacy rates and identify high- and low-performing states.

```{r}
ggplot(female_lit, aes(x = reorder(States, Female_Literacy),
                        y = Female_Literacy,
                       fill = Female_Literacy)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Rural Female Literacy by State",
       x = "State",
       y = "Literacy Rate")
```

The bar chart displays rural female literacy rates across states in ascending order.The visualization highlights clear disparities between states, indicating that rural female literacy is unevenly distributed across the country.

## Distribution of Key Variables

### Infant Mortality Rate

```{r}
ggplot(female_lit, aes(x = reorder(States, -Infant_Mortality_Rate),
                        y = Infant_Mortality_Rate)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +
  labs(title = "Infant Mortality Rate by State (Rural)",
       x = "States",
       y = "Deaths per 1000 Live Births")
```

### Fertility Rate

```{r}
ggplot(female_lit, aes(x = reorder(States,
                        `Total Fertility Rate (number of children per woman)`),
                        y = `Total Fertility Rate (number of children per woman)`)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(title = "Total Fertility Rate by State (Rural)",
       x = "State",
       y = "Children per Woman")
```

### Use of Birth Control Methods

```{r}
ggplot(female_lit, aes(x = reorder(States,
        `Use of Modern Family Planning Methods (%)`),
        y = `Use of Modern Family Planning Methods (%)`)) +
  geom_bar(stat = "identity", fill = "purple") +
  coord_flip() +
  labs(title = "Use of Modern Family Planning Methods (Rural)",
       x = "State",
       y = "Percentage (%)")
```

## Exploring the relationship between the variables

### Literacy vs Infant Mortality Rate

```{r}
#Literacy vs IMR
ggplot(female_lit, aes(x = Female_Literacy, 
                     y = Infant_Mortality_Rate)) +
  geom_point(aes(colour = Female_Literacy)) +
  geom_smooth(method = "lm", colour = "aquamarine2", se = FALSE) +
  labs(title = "Rural Female Literacy vs Infant Mortality Rate",
       x = "Rural Female Literacy",
       y = "Infant Mortality Rate")
```

A strong inverse relationship is observed between rural female literacy and infant mortality rate. States with higher literacy among women consistently report lower infant mortality, highlighting the critical role of education in improving child health outcomes. Educated women are more likely to access healthcare services, follow better nutrition and hygiene practices and ensure child vaccination.

### Literacy vs Total Fertility Rate

```{r}
#Literacy vs Fertility
ggplot(female_lit, aes(x = Female_Literacy, 
                      y = `Total Fertility Rate (number of children per woman)`)) +
  geom_point(aes(colour = Female_Literacy)) +
  geom_smooth(method = "lm", colour = "aquamarine2", se = FALSE) +
  labs(title = "Rural Female Literacy vs Fertility Rate",
       x = "Rural Female Literacy",
       y = "Fertility Rate")
```

Rural female literacy shows a moderate negative association with fertility rates, suggesting that increased education contributes to smaller family sizes and more controlled population growth. Educated women tend to delay marriage and childbirth and make informed reproductive choices.

### Literacy vs Use of Birth Control Methods

```{r}
#Literacy vs Birth Control Access

ggplot(female_lit, aes(x = Female_Literacy, 
                     y = `Use of Modern Family Planning Methods (%)`)) +
  geom_point(aes(colour = Female_Literacy)) +
  geom_smooth(method = "lm", colour = "aquamarine2", se = FALSE) +
  labs(title = "Rural Female Literacy vs Birth Control Access",
       x = "Rural Female Literacy",
       y = "Birth Control Access")
```

The relationship between rural female literacy and access to birth control appears weak and slightly negative, indicating that factors beyond education — such as cultural norms and public health outreach — play a significant role in determining contraceptive use.

Among the variables analyzed, infant mortality rate exhibits the strongest relationship with rural female literacy, followed by fertility rate, while access to birth control shows a comparatively weak and inconsistent association.

## Correlation Analysis
A correlation analysis was conducted to examine the relationships between rural female literacy and key demographic indicators.

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
install.packages("corrplot")
library(corrplot)
```

```{r}
literacy_cor <- female_lit %>%
  select(Female_Literacy, Infant_Mortality_Rate,
         `Total Fertility Rate (number of children per woman)`,
         `Use of Modern Family Planning Methods (%)`) %>%
  mutate(across(everything(), as.numeric))
```

```{r}
#Shortening col names for a cleaner map
colnames(literacy_cor) <- c("Literacy", "IMR", "Fertility", "BirthControl")

#running correlation
cor_matrix <- cor(literacy_cor, use = "complete.obs")
cor_matrix
```

Female literacy exhibits a moderately strong negative correlation with infant mortality rate (-0.56), indicating that higher levels of education among women are associated with improved child health outcomes and reduced infant deaths.

A weak negative correlation (-0.23) is observed between female literacy and fertility rate, suggesting that while education contributes to lower birth rates, its impact is limited compared to other influencing factors.

The relationship between female literacy and the use of modern family planning methods is weak and slightly negative (-0.24), suggesting that access to or use of contraceptive methods is influenced by factors beyond education alone.


```{r}
#Heatmap
corrplot(
  cor_matrix,
  method = "color",
  addCoef.col = "black",
  tl.cex = 0.9,
  number.cex = 0.8,
  tl.srt = 45
)
```

The correlation heat map highlights that female literacy is most strongly associated with reductions in infant mortality, while its relationship with fertility and contraceptive use remains relatively weak. In contrast, access to modern family planning methods exhibits a strong influence on fertility rates. These findings suggest that improving education and expanding healthcare access together are essential for achieving better demographic and health outcomes.

## Regression Analysis

Building on the correlation analysis, this section applies linear regression to evaluate the extent to which female literacy predicts variations in key socio-health indicators. This provides a deeper understanding of both strength and significance of these relationships.

```{r}
reg_data <- female_lit %>%
  select(`Total Fertility Rate (number of children per woman)`, Female_Literacy, Infant_Mortality_Rate,`Use of Modern Family Planning Methods (%)`) %>%
  na.omit()

#Infant Mortality Rate
model_imr <- lm(Infant_Mortality_Rate ~ Female_Literacy, data = reg_data)

#Fertility Rate
model_tfr <- lm(`Total Fertility Rate (number of children per woman)` ~ Female_Literacy, data = reg_data)

#Birth Control
model_bc <- lm(`Use of Modern Family Planning Methods (%)` ~ Female_Literacy, data = reg_data)
```
```{r}
summary_table <- data.frame(
  Variable = c("Infant Mortality Rate", "Fertility Rate", "Birth Control Use"),
  Intercept = c(coef(model_imr)[1], coef(model_tfr)[1], coef(model_bc)[1]),
  Slope = c(coef(model_imr)[2], coef(model_tfr)[2], coef(model_bc)[2]),
  R_squared = c(summary(model_imr)$r.squared,
                summary(model_tfr)$r.squared,
                summary(model_bc)$r.squared),
  P_value = c(summary(model_imr)$coefficients[2,4],
              summary(model_tfr)$coefficients[2,4],
              summary(model_bc)$coefficients[2,4])
)

library(kableExtra)

summary_table %>%
  kable(digits = 3, caption = "Regression Results Summary") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

### Infant Mortality Rate

There is a statistically significant negative relationship between female literacy and infant mortality rate. Specifically, a 1 percentage point increase in female literacy is associated with a decrease of approximately 0.58 infant deaths per 1000 live births. The model explains about 31% of the variation in infant mortality across states, indicating that while literacy is an important determinant, other socio-economic and healthcare factors also play a role. 

### Fertility Rate

The regression analysis shows a negative but statistically insignificant relationship. Although higher literacy is associated with a slight decline in fertility, the effect is minimal and not statistically meaningful (p > 0.05). The model explains only about 5% of the variation in fertility rates, indicating that female literacy alone is not a strong predictor of fertility in rural India.

### Birth Control

The regression analysis shows a negative but statistically insignificant relationship. Although the coefficient suggests a decline in contraceptive use with increasing literacy, this result is not statistically meaningful (p > 0.05). The model explains only about 6% of the variation, indicating that literacy alone is not a strong determinant of contraceptive usage.

## Conclusion

This analysis highlights important relationships between female literacy, infant mortality, fertility, and birth control usage.

Female literacy plays a key role in improving child health outcomes, as seen in its strong negative correlation with infant mortality. However, literacy alone has a limited direct impact on fertility reduction. The most influential factor in lowering fertility is access to modern birth control methods, which shows the strongest relationship in the dataset.

Governments should prioritize universal female education as a long-term strategy for improving public health. Availability and awareness of modern contraceptive methods should be increased, especially in rural and underserved areas. Further investment should be directed towards prenatal care, immunization, and nutrition programs.

A combined policy approach is essential for sustainable demographic and public health improvements.




#### The data used in this analysis is sourced from: 
National Family Health Survey (NFHS-5), India, (https://www.data.gov.in/resource/india-districts-factsheets-national-family-health-survey-nfhs-5-2019-2021-provisional)










